---
title: "CMSC320_Final"
author: "Michael Shultz"
date: "May 8, 2018"
output: html_document
---
#Honey Production in the United States of America

###Background
Honeybees are an integral part of the ecosystem. They help pollinate flowers and other plants, and are essential to food growth in the United States. [This](http://www.lifebasicsorganics.com/blog/why-are-bees-important) article gives some interesting insight and more detail into what honeybees do for both the environment and human food supply. 
One of the greatest parts about honeybees, however, is the fact that they make honey. According to the [USDA National Honey Report](https://www.ams.usda.gov/mnreports/fvmhoney.pdf), the U.S. had exported 657,813 kg of honey in 2018 as of April 24. I aim to use previous years to predict honey production for upcoming years.  

###Getting our data
```{r load_package, message=FALSE, warning=FALSE}
library("tidyr", lib.loc="~/R/win-library/3.4") 
library("dplyr", lib.loc="~/R/win-library/3.4")
library("readr", lib.loc="~/R/win-library/3.4")
library("tidyverse", lib.loc="~/R/win-library/3.4")
# library("rJava", lib.loc="~/R/win-library/3.4")
# library("xlsx", lib.loc="~/R/win-library/3.4")
```
With our packages loaded, let's read in our data and take a look at what we have. The dataset comes from [this](https://www.kaggle.com/jessicali9530/honey-production/version/2) kaggle page.
```{r data, message=FALSE}
honey_df <- read_csv("honeyproduction.csv") #Data from site
agr_df <- read_csv("table03.xls.csv") #Read state agricultural data
honey_df %>% head() 
agr_df %>% head()
```

Looks like we have some cleaning to do for our agricultural data set. A look at the middle rows will show us that we have 
```{r view2}
agr_df %>% slice(50:60)
```
So it looks like our cleaning is going to involve splitting this data set, getting better labels, and switching our rows and columns.

# agr_df <- agr_df %>%
#   slice(c(5,8:100)) 
colnames(agr_df) = agr_df[5, ]
agr_df <- agr_df %>%
  slice(8:100)
```
Here is a basic explanation of the attributes present in the dataset:

Attribute    | Description                                   
------------ | -------------------------------------------------------
state        | State data is taken from
numcol       | Colonies from which honey was taken that year 
yieldpercol  | Honey yield per colony in pounds              
totalprod    | Total honey production (numcol x yieldpercol) in pounds
stocks       | Stocks held by producers in pounds
priceperlb   | Average price per pound in dollars
prodvalue    | Value of production (totalprod x priceperlb) in dollars
year         | Year data is taken from

