---
title: "CMSC320_Final"
author: "Michael Shultz"
date: "May 8, 2018"
output: html_document
---
```{r setup, include=FALSE}
library("tidyr", lib.loc="~/R/win-library/3.4")
library("dplyr", lib.loc="~/R/win-library/3.4")
library("readr", lib.loc="~/R/win-library/3.4")
library("tidyverse", lib.loc="~/R/win-library/3.4")
```
#Honey Production in the United States of America

###Background
Honeybees are an integral part of the ecosystem. They help pollinate flowers and other plants, and are essential to food growth in the United States. [This](http://www.lifebasicsorganics.com/blog/why-are-bees-important) article gives some interesting insight and more detail into what honeybees do for both the environment and human food supply. 
One of the greatest parts about honeybees, however, is the fact that they make honey. According to the [USDA National Honey Report](https://www.ams.usda.gov/mnreports/fvmhoney.pdf), the U.S. had exported 657,813 kg of honey in 2018 as of April 24. In 2006, [colony collapse disorder](http://www.abc.net.au/news/2017-05-08/colony-collapse-ten-years-after-crisis-what-is-happening-to-bees/8507408) crippled the backbone of American honey production. This dataset tracks honey production and sales in the United States from 1998 to 2012.

###Getting our data
The dataset comes from [this](https://www.kaggle.com/jessicali9530/honey-production/version/2) kaggle page. 
```{r data, message=FALSE}
honey_df <- read_csv("honeyproduction.csv")

honey_df %>% head()
```